The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-05-04 paper/ms.tex
Fix lead-ins to equations

Once a Galactocentric coordinate system has been established and a
six-dimensional (6D) phase space measurement of a star has been made, it is
often desirable to convert this measurement into action space to concisely
summarize its projected orbit, model the stellar distribution function, or
find stars with similar dynamical properties. Actions are conserved quantities
that describe the orbit of a star under the assumption of regular, bound
orbits in a system where the equations of motion are separable in a particular
coordinate system. They are the cyclical integral of the canonical momentum
over its conjugate position,
\beq\label{eq:actions}
J_i \equiv
\frac{1}{2\pi} \oint_{\text{orbit}}p_i\,\text{d}x_i\text{.}
\eeq
Under the assumption of axisymmetry, $i=R,z,\phi$ are the radial, vertical,
and azimuthal coordinates respectively in a cylindrical coordinate system and
the $p_i$ are the conjugate momenta. In a slowly-evolving axisymmetric
potential, these actions are invariant and $J_{\phi} \equiv L_z$, where $L_z$ is the $z$-component of the angular momentum
\citep{2008gady.book.....B,2014RvMP...86....1S}.

@@

Under this approximation, we can write down the cylindrical components
of the orbits as,
\beq\label{eq:orbits_epi}
\begin{split}
\phi(t) &= \Omega_c t \\
R(t) &= R_g + A_R \sin{(\kappa t + \alpha)} \\
z(t) &= A_z \sin{(\nu t + \beta)}
\text{,}
\end{split}
\eeq
where $\kappa$ and $\nu$ are the radial and vertical frequencies, $\Omega_c
\equiv v_c/R_g$ is the rotation frequency, $A_z$ and $A_R$ are the amplitudes of
the stellar motion in the vertical and radial coordinates, and $\alpha$ and
$\beta$ are arbitrary constants. Similarly, the velocities of the orbit are
given by
\beq\label{eq:orbits_vel_epi}
\begin{split}
v_{\phi}(t) &= v_c \\
v_R(t) &= \kappa A_R \cos{(\kappa t + \alpha)} \\
v_z(t) &= \nu A_z \cos{(\nu t + \beta)}
\text{.}
\end{split}
\eeq

In this case, we have that for $J_{\phi}$
\citep[][Section~3.5.3b]{2008gady.book.....B},
\beq\label{eq:Jphi_epi}
J_{\phi} = R_g v_c\text{,}
\eeq
and for $J_R$ and $J_z$,
\beq\label{eq:JR_Jz_epi}
\begin{split}
J_R &= \frac{E_R}{\kappa} \\
J_z &= \frac{E_z}{\nu} \text{,}
\end{split}
\eeq
where $E_R$ and $E_z$ are the energy per unit mass in the radial and vertical
coordinates, respectively. Therefore,
\beq\label{eq:JR_Jz_epi_energy}
\begin{split}
J_R &= \frac{v_R^2 + \kappa^2 (R-R_g)^2}{2\kappa} \\
J_z &= \frac{v_z^2 + \nu^2 z^2}{2\nu}\text{.}
\end{split}
\eeq
We can simplify this and write,
\beq\label{eq:JR_Jz_epi_final}
\begin{split}
J_R &= \frac{\kappa A_R^2}{2} = \frac{v_{R,\text{max}}^2}{2\kappa} \\
J_z &= \frac{\nu A_z^2}{2} = \frac{v_{z,\text{max}}^2}{2\nu}\text{,}
\end{split}
\eeq
where the last equality comes from the fact that $v_{R,\text{max}} = \kappa
A_R$ and similarly for $z$.

Notice that while the value for each of $J_{\phi}$, $J_R$, and $J_z$ is phase
independent, the contribution from the kinetic and potential terms in
Equation~\eqref{eq:JR_Jz_epi_energy} is phase dependent. Now assume that the
coordinates $(R, z, v_{\phi}, v_R, v_z)$ are offset by $(\Delta R, \Delta z,
\Delta v_{\phi}, \Delta v_R, \Delta v_z)$. We can then apply the standard
propagation of errors formula to Equation~\eqref{eq:JR_Jz_epi_energy}
to determine the error in each of the actions. For $J_{\phi}$ we have that,
\beq\label{eq:induced_Jphi}
\frac{\Delta J_{\phi}}{J_{\phi}} = \frac{\Delta R}{R_g} 
                                    + \frac{\Delta v_{\phi}}{v_c}\text{.}
\eeq
For $J_R$, the induced error is,
\beq\label{eq:induced_JR}
\frac{\Delta J_R}{J_R} = \frac{2(R-R_g)}{A_R^2}\Delta R
                         + \frac{2v_R}{v_{R,\text{max}}^2} \Delta v_R \text{.}
\eeq
For $J_z$, the induced error is,
\beq\label{eq:induced_Jz}
\frac{\Delta J_z}{J_z} = \frac{2z}{A_z^2}\Delta z
                         + \frac{2v_z}{v_{z,\text{max}}^2} \Delta v_z \text{.}
\eeq
We have ignored second order contributions.

>>>>

Once a Galactocentric coordinate system has been established and a
six-dimensional (6D) phase space measurement of a star has been made, it is
often desirable to convert this measurement into action space to concisely
summarize its projected orbit, model the stellar distribution function, or
find stars with similar dynamical properties. Actions are conserved quantities
that describe the orbit of a star under the assumption of regular, bound
orbits in a system where the equations of motion are separable in a particular
coordinate system. They are the cyclical integral of the canonical momentum
over its conjugate position:
\beq\label{eq:actions}
J_i \equiv
\frac{1}{2\pi} \oint_{\text{orbit}}p_i\,\text{d}x_i\text{.}
\eeq
Under the assumption of axisymmetry, $i=R,z,\phi$ are the radial, vertical,
and azimuthal coordinates respectively in a cylindrical coordinate system and
the $p_i$ are the conjugate momenta. In a slowly-evolving axisymmetric
potential, these actions are invariant and $J_{\phi} \equiv L_z$, where $L_z$ is the $z$-component of the angular momentum
\citep{2008gady.book.....B,2014RvMP...86....1S}.

@@

Under this approximation, we can write down the cylindrical components
of the orbits as
\beq\label{eq:orbits_epi}
\begin{split}
\phi(t) &= \Omega_c t \\
R(t) &= R_g + A_R \sin{(\kappa t + \alpha)} \\
z(t) &= A_z \sin{(\nu t + \beta)}
\text{,}
\end{split}
\eeq
where $\kappa$ and $\nu$ are the radial and vertical frequencies, $\Omega_c
\equiv v_c/R_g$ is the rotation frequency, $A_z$ and $A_R$ are the amplitudes of
the stellar motion in the vertical and radial coordinates, and $\alpha$ and
$\beta$ are arbitrary constants. Similarly, the velocities of the orbit are
given by:
\beq\label{eq:orbits_vel_epi}
\begin{split}
v_{\phi}(t) &= v_c \\
v_R(t) &= \kappa A_R \cos{(\kappa t + \alpha)} \\
v_z(t) &= \nu A_z \cos{(\nu t + \beta)}
\text{.}
\end{split}
\eeq

In this case, we have that for $J_{\phi}$
\citep[][Section~3.5.3b]{2008gady.book.....B}:
\beq\label{eq:Jphi_epi}
J_{\phi} = R_g v_c\text{,}
\eeq
and for $J_R$ and $J_z$:
\beq\label{eq:JR_Jz_epi}
\begin{split}
J_R &= \frac{E_R}{\kappa} \\
J_z &= \frac{E_z}{\nu} \text{,}
\end{split}
\eeq
where $E_R$ and $E_z$ are the energy per unit mass in the radial and vertical
coordinates, respectively. Therefore,
\beq\label{eq:JR_Jz_epi_energy}
\begin{split}
J_R &= \frac{v_R^2 + \kappa^2 (R-R_g)^2}{2\kappa} \\
J_z &= \frac{v_z^2 + \nu^2 z^2}{2\nu}\text{.}
\end{split}
\eeq
We can simplify this and write
\beq\label{eq:JR_Jz_epi_final}
\begin{split}
J_R &= \frac{\kappa A_R^2}{2} = \frac{v_{R,\text{max}}^2}{2\kappa} \\
J_z &= \frac{\nu A_z^2}{2} = \frac{v_{z,\text{max}}^2}{2\nu}\text{,}
\end{split}
\eeq
where the last equality comes from the fact that $v_{R,\text{max}} = \kappa
A_R$ and similarly for $z$.

Notice that while the value for each of $J_{\phi}$, $J_R$, and $J_z$ is phase
independent, the contribution from the kinetic and potential terms in
Equation~\eqref{eq:JR_Jz_epi_energy} is phase dependent. Now assume that the
coordinates $(R, z, v_{\phi}, v_R, v_z)$ are offset by $(\Delta R, \Delta z,
\Delta v_{\phi}, \Delta v_R, \Delta v_z)$. We can then apply the standard
propagation of errors formula to Equation~\eqref{eq:JR_Jz_epi_energy}
to determine the error in each of the actions. For $J_{\phi}$, the induced error is:
\beq\label{eq:induced_Jphi}
\frac{\Delta J_{\phi}}{J_{\phi}} = \frac{\Delta R}{R_g} 
                                    + \frac{\Delta v_{\phi}}{v_c}\text{.}
\eeq
For $J_R$, the induced error is:
\beq\label{eq:induced_JR}
\frac{\Delta J_R}{J_R} = \frac{2(R-R_g)}{A_R^2}\Delta R
                         + \frac{2v_R}{v_{R,\text{max}}^2} \Delta v_R \text{.}
\eeq
For $J_z$, the induced error is:
\beq\label{eq:induced_Jz}
\frac{\Delta J_z}{J_z} = \frac{2z}{A_z^2}\Delta z
                         + \frac{2v_z}{v_{z,\text{max}}^2} \Delta v_z \text{.}
\eeq
We have ignored second order contributions.
